---
output:
  github_document:
    html_preview: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 3.5,
                      fig.width = 5, fig.align = 'center',
                      warning = FALSE, message = FALSE,
                      cache = TRUE)
library(rebird) #get eBird data from the web
library(dplyr) #data munging
library(purrr) #map() for looping
library(ggplot2) #plotting
library(stringr) #clean you variables
library(lubridate) #make date data dates
library(tidyr) #separate month and quarter
library(rphylopic) #for image, package from GitHub
library(curl) #required by rphylopic function
```

# Where are the Rufus Hummingbirds?

<img src = "../images/vnhs_ruhu_tweet.png" width = "500"></img>

When I saw the above analysis by my friend [Dave Fraser](https://twitter.com/DavidFFraser), I thought `r emo::ji("thought")`:

- cool analysis!
- citizen science + [eBird](https://ebird.org/science/download-ebird-data-products) is awesome!
- have I seen a Rufus Hummingbird lately? `r emo::ji("bird")`

_and then I thought ..._  

- maybe I could do that in R? `r emo::ji("idea")`

And then I did.

## Getting the [eBird](https://ebird.org/science/download-ebird-data-products) Data

You can get up-to-date eBird data&mdash;once you have created an accout and logged in&mdash;via the [eBird website GUI](https://ebird.org/explore), but I was hoping to include more years and didn't want to click that often. The [full eBird data set is also available for download](https://ebird.org/science/download-ebird-data-products), but it is updated quarterly&mdash;so is unlikely to have the very recent records I am after. Sigh `r emo::ji("disappointed")`.

And then I talked to [Andy Teucher](https://github.com/ateucher) about my data woes. He introduced me to the [rOpenSci](https://ropensci.org/) [R](https://www.r-project.org/) package `r emo::ji("package")` [`rebird`](https://cran.r-project.org/web/packages/rebird/index.html)&mdash;he is a coauthor! `r emo::ji("+1")`

The [`rebird`](https://cran.r-project.org/web/packages/rebird/index.html) package `r emo::ji("package")` gets the eBird data from the web and provides it in a tidy format. `r emo::ji("present")`

So let's start by pulling _A LOT_ of eBird data&mdash;50 years seems like a lot?

```{r get-data}
#load R packages we will need 
# library(rebird) #get eBird data from the web using ebirdfreq()
# library(dplyr) #data munging
# library(purrr) #map() for looping
# 
# #function to get eBird data for many states/provinces using rebird R package
# map_state <-  function(state) {
#   map_dfr(1968:2018, ~ {
#     ebirdfreq("states", state, .x, .x, 1, 5) %>%
#       filter(comName == "Rufous Hummingbird") %>%
#       mutate(year = .x, state = state)
#   })
# }
# 
# #get eBird data for west coast state/provinces
# ruhu_raw <- map_dfr(c("CA-BC","US-CA","US-WA", "US-OR"), ~ {
#       map_state(.x)
# })
# 
# #save data object in tmp folder
# if (!exists("tmp")) dir.create("tmp", showWarnings = FALSE)
# save(ruhu_raw, file = "tmp/ruhu_raw.RData")

```

```{r}
# temp chunk to load saved raw data file
if (!exists("ruhu_raw")) load("tmp/ruhu_raw.RData")
```


## Some Data Cleaning 

```{r}
head(ruhu_raw)
```


The eBird data provided by [`rebird`](https://cran.r-project.org/web/packages/rebird/index.html) is indeed very tidy, perfect! `r emo::ji("present")` Let's take a moment to cheers `r emo::ji("tada")` the authors of this open source R package. OK, now let's keep going... 

We need to do something with the `monthQt` variable. This is our time variable, which we need for plotting, and it needs to be a recognized as a date. And you might notice some zeros in the 2018 weeks that we have not yet lived&mdash;or birded. Let's filter those zeros, that should be `NA`s, out of the data frame.

```{r clean-data}
#library(dplyr) #data munging & already loaded
library(stringr) #clean up variables
library(lubridate) #make date data dates
library(tidyr) #separate month and quarter

ruhu_clean <- ruhu_raw %>% 
  separate(monthQt, c("month", "week"), sep = "-") %>%
   mutate(day = case_when(
    week == 1 ~ 1,
    week == 2 ~ 8,
    week == 3 ~ 15,
    week == 4 ~ 23
  )) %>%
  mutate(date = ymd(paste0(year, "-",month,"-",day))) %>% 
  mutate(week = lubridate::week(date)) %>% 
  filter(sampleSize != 0) #filter out any zero that is actually an NA

head(ruhu_clean)
```

That looks `r emo::ji("eyes")` better!

## Some Data Visualizing

Now let's have a quick, exploratory look `r emo::ji("eyes")` at what data we have, starting just with British Columbia.

```{r}
library(ggplot2) #plotting

ruhu_clean %>% 
  filter(state == "CA-BC") %>%  #just B.C.
ggplot(aes(x = week, y = frequency, group = year)) +
  geom_line(colour = "grey") +
  theme_minimal()

```

There seems to be A LOT of very high frequency values in the data set (frequency = 1.0), which is a bit puzzling? Let's colour the lines for each year to see if we learn something.

```{r}

ruhu_clean %>% 
  filter(state == "CA-BC") %>%  #just B.C.
ggplot(aes(x = week, y = frequency, group = year)) +
  geom_line(aes(colour = year)) +
  theme_minimal()

```

Hmmmmm.. `r emo::ji("eyes")` This shows that most of that puzzling data is OLDER, pre-2000s. Andy&mdash;a birder and frequent data provider to eBird&mdash;mentioned that the older records are being entered by birders going through old checklists & field notebooks. Perhaps this historical data is not as complete as data since 2002&mdash; the year eBird launched and birders started entering sightings via smartphones? Filtering out the pre-2002 frequency data seems like a reasonable approach to me.

```{r filter}

ruhu_since_2002 <- ruhu_clean %>% 
  filter(year > 2001) 

ruhu_since_2002 %>% 
  filter(state == "CA-BC") %>%  #just B.C.
ggplot(aes(x = week, y = frequency, group = year)) +
  geom_line(aes(colour = year)) +
  theme_minimal()
```

Looks `r emo::ji("eyes")` better! 

## How Does 2018 `r emo::ji("bird")` Observation Frequencies Compare?

We want to compare 2018 observation frequency against previous years. 
